---
title: "1. Data Extraction"
author: "Anu Joshi"
date: "May 16, 2019"
output:
pdf_document
    html_document:
    keep_md: yes
    theme: united
    highlight: tango
---

```{r setup,  message=FALSE}
knitr::opts_chunk$set(echo = TRUE, 
                      warning = FALSE)

# Load libraries
library(readxl)                 # read Excel files
library(tidyverse)              # data manipulation
library(magrittr)               # data manipulation
```

```{r load-metabolite-dats, results='hide', message=FALSE}
metabolite = read_excel("data/MSSM-01-18ML+ CLIENT DATA TABLE 6.14.19.xlsx", 
                     sheet = "OrigScale", 
                     col_names = FALSE)
```
<br>


**1. Extract metabolite identifiers** 
```{r extract-id}
metabolite_IDs = metabolite[14:1321, 1:13] %>%
  set_colnames(metabolite[13, 1:13])

write_csv(metabolite_IDs, "data/1_metabolite-details.csv")
```
<br>


**2. Extract metabolite concentrations**     
In genetic datasets, the general convention is to have the participants IDs in rows and the metabolites in columns when the number of variables are greater than the number of participants (n > p).  
<br> 

Using Assay ID as participant identifier (in rows) and COMP ID as metabolite identifiers (in columns). Of all the metabolite identifiers, "COMP ID" has the most complete data so we are going to use it to extract our metabolite information. It is an arbitary value assigned for identifying metabolites. 
```{r extract-compID}
# transpose the information because this is going to be used as column names
compID = metabolite[c(14:1321), 5] %>% 
    t() %>%
    paste0("C",.)


df_compID = metabolite[c(3, 14:1321), 14:513] %>%
    t() %>%
    as_tibble() %>%
    set_colnames(compID)
colnames(df_compID)[1] = "Assay"

write_csv(df_compID, "data/1_metabolite.csv")
```
We can match Assay ID with the studyID from the covariate data to merge and create a complete participant history file.  
<br>
