---
title: "1. Data Extraction"
author: "Anu Joshi"
date: "May 16, 2019"
output:
    html_document:
    keep_md: yes
    theme: united
    highlight: tango
---

```{r setup, message=FALSE}
knitr::opts_chunk$set(echo = TRUE, 
                      message = FALSE, 
                      warning = FALSE, 
                      comment = "", 
                      tidy = TRUE)

# Load libraries
library(here)                   # for universal pathways
library(janitor)                # for clean names
library(readxl)                 # read Excel files
library(tidyverse)              # data manipulation
library(magrittr)               # data manipulation
```

```{r include=FALSE}
metabolite <- read_excel(here("data","MSSM-01-18ML+ CLIENT DATA TABLE 6.14.19.xlsx"),
                         sheet = "OrigScale", 
                         col_names = FALSE)
```
<br>


**1. Extract metabolite identifiers** 

This metabolite dataset has metadata in the first 13 columns. Extract that and save it separately.

```{r extract-id}
metabolite_IDs <- 
  metabolite[14:1321, 1:13] %>%
  set_colnames(metabolite[13, 1:13]) %>% 
  janitor::clean_names()

write_csv(metabolite_IDs, here("data", "1_metabolite-details.csv"))
```
<br>


**2. Extract metabolite concentrations**     

In genetic dataset, the general convention is to have the participants IDs in rows and the metabolites in columns when the number of variables are greater than the number of participants (n > p).  
<br> 

Using `Assay ID` as participant identifier (in rows) and `COMP ID` as metabolite identifiers (in columns). Of all the metabolite identifiers, `COMP ID` has the most complete data so we are going to use it to extract our metabolite information. It is an arbitrary value assigned for identifying metabolites. 

```{r extract-compID}
# transpose the information because this is going to be used as column names
compID <- 
  metabolite[c(14:1321), 5] %>% 
  t() %>%
  paste0("C",.)


df_compID <- 
  metabolite[c(3, 14:1321), 14:513] %>%
  t() %>%
  as_tibble() %>%
  set_colnames(compID)
colnames(df_compID)[1] = "Assay"

write_csv(df_compID, here("data", "1_metabolite.csv"))
```

We can match Assay ID with the studyID from the covariate data to merge and create a complete participant history file.  
<br>
