---
title: "2. Data Cleaning"
author: "Anu Joshi"
date: "5/17/2019"
output:
  html_document:
    keep_md: yes
    theme: united
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, 
                      warning = FALSE)
```

```{r load-libraries, results='hide', message=FALSE}
library(readxl)                 # to read Excel files
library(dplyr)                  # for data manipulation
library(magrittr)               # for data manipulation
library(tibble)                 # for row and data manipulation
```

```{r load-metabolite-dats, results='hide', message=FALSE}
metabolite = read_excel("data/MSSM-01-18ML+ CLIENT DATA TABLE 9.25.18.XLSX", 
                     sheet = "OrigScale", 
                     col_names = FALSE)
```
<br>

There are multiple identifiers associated with each metabolite. Here we extract all the identifiers for latter use, as all of them are not required for the analysis. 
```{r extract-id}
metabolite_IDs = metabolite[10:1317, 1:13] %>%
  set_colnames(metabolite[9, 1:13])

# quick view of the identifiers
head(metabolite_IDs)
```
<br>


**Extract the table of metabolite concentrations**     

Of all the metabolite identifiers we have on hand, "_COMP ID_" has the most complete data so we are going to use it to extract our metabolite information. COMP-ID is an arbitary value assigned by Metabolon for metabolites identification.
```{r extract-compID}
# transpose the information because this is going to be used as column names
compID = metabolite[c(9:1317), 5] %>% t()
```


We are using the Assay ID as participant identifier, this will be later matched with studyID from the covariate data to merge and create a complete file.    
In genetic datasets the participants IDs are in rows and the metabolites are in columns. So, we tranpose our dataset to follow that convention. This is the general rule when the number of variables are greater than the number of participants (n > p).  
```{r extract-metabolite}
df_compID = metabolite[c(3, 10:1317), 14:513] %>%
    t() %>%
    as_tibble() %>%
    set_colnames(compID)
colnames(df_compID)[1] = "Assay"
```
<br>

**Remove metabolites with >30% missing data**

We want to remove a metabolite if it has more than 30% missing value, because it would affect the data quality when we do data imputation. 
```{r subset}
# remove Assay ID
# make sure all the data is numeric, in order to count the missing values
df <- df_compID[, -1] %>%
    mutate_if(is.character, as.numeric) 

# select a threshold value: 30% of the total metabolites
# the number of missing values for each metabolite should be lower than the 30% threshold
n = ncol(df)*0.30

# subset data with less than 30% missing values
# removed 93 metabolites
df_clean = subset(df, select = colSums(is.na(df))< n) %>%
    cbind(Assay = df_compID$Assay, .)
```
<br>


Add metabolite identifiers to the 93 metabolites that we are looking to drop.        
_Note_: Most of the metabolites with $>30%$ missing value are Xenobiotics. If you are interested in those pathway, you can choose to keep them. It is important to explore the data before dropping them. 
```{r dropped-metabolites}
missing30 = subset(df, select = colSums(is.na(df)) > n) %>% colnames()

id_missing30 <- match(missing30, metabolite_IDs$`COMP ID`)
df_missing30 = metabolite_IDs[id_missing30, ]
```
<br>

```{r save-files}
write.csv(metabolite_IDs, "data/1_metabolite-details.csv", row.names = FALSE)
write.csv(df_compID, "data/1_metabolites.csv", row.names = FALSE)
write.csv(df_missing30, "data/1_removed-metabolites.csv", row.names = FALSE)
write.csv(df_clean, "data/1_metabolites-clean.csv", row.names = FALSE)
```