---
title: "2. Data Cleaning"
author: "Anu Joshi"
date: "May 17, 2019"
output:
  html_document:
    keep_md: yes
    theme: united
---

```{r setup, message=FALSE}
knitr::opts_chunk$set(echo = TRUE, 
                      message = FALSE, 
                      warning = FALSE, 
                      comment = "", 
                      tidy = TRUE)

# Load libraries
library(here)                   # for universal pathways
library(janitor)                # for clean names
library(readr)                  # faster reading and writing csv files
library(dplyr)                  # data manipulation
library(magrittr)               # pipe operator
```
<br>

## 500 observations of 1314 variables
```{r load-data, message=FALSE}
df = read_csv(here("data", "1_metabolite.csv")) 
```


1. Remove individuals that had their blood drawn incorrectly or have withdrawn/been removed from the study
## Removed 30 individuals
## 470 observations of 1314 variables
```{r errors}
# step 1: load file of individuals that have been marked for second review
remove = read_csv("data/outliers.csv")

# step 2: select individuals that have been removed from the main data
remove0 = subset(remove, remove$Exclude == 1) %>% select(Assay)

# step 3: remove the erroneous Assay ID
df_clean = df[!df$Assay %in% remove0$Assay, ]

# step 4: remove unnecessary files
rm(remove, remove0)
```


2. Add household ID and individual ID information for easy identification
## Removed 11 individuals with duplicate ID
## 459 observations of 1312 variables
```{r studyID}
# step 1:  extract necessary variables from the ID file
id = read_csv("data/PRISM linking IDs_clean.csv")[, c("Assay", "type", "HID", "IID_2", "Cohort")]

# step 2: change the nomenclature to reflect the file to match
id$Assay = if_else(id$type == "original", paste0(id$Assay, " M"), paste0(id$Assay, " A2"))

# step 3: merge the ID file with the full data file, keep all the metabolite data
# 470 observations of 1313 variables
df_clean_id = merge(id, df_clean, by = "Assay", all.y = TRUE)

# step 4: remove duplicate ID
# 459 observations of 1312 variables
# Removed 11 duplicated IDs
df_clean_id0 = subset(df_clean_id, df_clean_id$type == "original") %>%
    select(-type)

# step 4: remove unnecessary files
rm(id, df_clean_id)
```


3. Create a new studyID variable.     
## There are a few household ID (HID) that have more than one child in the study, so HID cannot be used as an unique identifier. Concatenate HID-IID to create a new variable: studyID_2
## 459 observations of 1313 variables
```{r create-new-var}
# create a duplicate file 
clean <- df_clean_id0

# add a new variable, studyID_2
clean0 = clean %>%
    mutate(studyID_2 = paste0(clean$HID, "-", clean$IID_2)) %>%
    select(Assay, studyID_2, everything())
```


4. Add serum information 
## 459 observations of 1314 variables
```{r serum}
# step 1: load serum data
serum = read_csv("data/serum_wk.csv")

# step 2: create a new unique variable, studyID_2 = HID-IID
serum$iid[which(serum$hhid > "7000")] = serum$iid[which(serum$hhid > "7000")]*10
serum0 = serum %>%
    mutate(studyID_2 = paste0(serum$hhid, "-", serum$iid), .) %>%
    select(-hhid, -iid)


# step 3: add serum information to the main file
clean_serum = merge(serum0, clean0, by = "studyID_2", all.y = TRUE) %>%
    select(Assay, studyID_2, serum_wk, everything())
```


5. Save file
```{r save-output}
write_csv(clean_serum, "data/2_metabolite-clean.csv")
```